---
output: github_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nflfastR)
library(ggimage)
library(qs)
```



```{r load-data, include=FALSE, message=FALSE, echo=FALSE}
pbp <- read_csv("Data/play_by_play.csv")

max_week <- max(pbp$week)
```

# NFL Landscape: Week `r max_week`

### Overall Efficiency Landscape
The code for this graph (and data for the whole report) comes directly from the good people at [nflfastR](https://www.nflfastr.com/index.html) straight off their [getting started page](https://www.nflfastr.com/articles/nflfastR.html#example-5-plot-offensive-and-defensive-epa-per-play-for-a-given-season), and for that, I thank them.  

This graph looks at the offense vs defense epa for the current season as a whole, for every game. 
```{r season-efficiency, echo=FALSE, warning=FALSE}
pbp <-  pbp %>%
  dplyr::filter(season_type == "REG") %>%
  dplyr::filter(!is.na(posteam) & (rush == 1 | pass == 1))
offense <- pbp %>%
  dplyr::group_by(posteam) %>%
  dplyr::summarise(off_epa = mean(epa, na.rm = TRUE))
defense <- pbp %>%
  dplyr::group_by(defteam) %>%
  dplyr::summarise(def_epa = mean(epa, na.rm = TRUE))
logos <- teams_colors_logos %>% dplyr::select(team_abbr, team_logo_espn)
joined <- offense %>%
  dplyr::inner_join(defense, by = c("posteam" = "defteam")) %>%
  dplyr::inner_join(logos, by = c("posteam" = "team_abbr")) %>%
  dplyr::inner_join(teams_colors_logos, by = c("posteam" = "team_abbr"))

joined %>%
  ggplot2::ggplot(aes(x = off_epa, y = def_epa)) +
  ggplot2::geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
  ggplot2::geom_hline(aes(yintercept = mean(off_epa)), color = "red", linetype = "dashed") +
  ggplot2::geom_vline(aes(xintercept = mean(def_epa)), color = "red", linetype = "dashed") +
  ggimage::geom_image(aes(image = team_logo_espn.x), size = 0.05, asp = 16 / 9) +
  ggplot2::labs(
    x = "Offense EPA/play",
    y = "Defense EPA/play",
    caption = "Data: @nflfastR",
    title = "NFL Offensive and Defensive EPA per Play (Total Season)"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    plot.title = ggplot2::element_text(size = 12, hjust = 0.5, face = "bold")
  ) +
  ggplot2::scale_y_reverse()
```

But how is a team trending? Let's look at the same plot of offense & defense efficiency the last two games only. 

```{r recent-efficiency, echo=FALSE, warning=FALSE, message=FALSE}
pbp_recent <- pbp %>% 
  dplyr::filter(week + 1 >= max_week)

offense <- pbp_recent %>%
  dplyr::group_by(posteam) %>%
  dplyr::summarise(off_epa = mean(epa, na.rm = TRUE))
defense <- pbp_recent %>%
  dplyr::group_by(defteam) %>%
  dplyr::summarise(def_epa = mean(epa, na.rm = TRUE))
logos <- teams_colors_logos %>% dplyr::select(team_abbr, team_logo_espn)
joined <- offense %>%
  dplyr::inner_join(defense, by = c("posteam" = "defteam")) %>%
  dplyr::inner_join(logos, by = c("posteam" = "team_abbr")) %>%
  dplyr::inner_join(teams_colors_logos, by = c("posteam" = "team_abbr"))

joined %>%
  ggplot2::ggplot(aes(x = off_epa, y = def_epa)) +
  ggplot2::geom_abline(slope = -1.5, intercept = c(.4, .3, .2, .1, 0, -.1, -.2, -.3), alpha = .2) +
  ggplot2::geom_hline(aes(yintercept = mean(off_epa)), color = "red", linetype = "dashed") +
  ggplot2::geom_vline(aes(xintercept = mean(def_epa)), color = "red", linetype = "dashed") +
  ggimage::geom_image(aes(image = team_logo_espn.x), size = 0.05, asp = 16 / 9) +
  ggplot2::labs(
    x = "Offense EPA/play",
    y = "Defense EPA/play",
    caption = "Data: @nflfastR",
    title = "NFL Offensive and Defensive EPA per Play (Last 2 games)"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    plot.title = ggplot2::element_text(size = 12, hjust = 0.5, face = "bold")
  ) +
  ggplot2::scale_y_reverse()
```

## Offense and Defensive Breakdowns 

Where are these efficiency metrics coming from and what drives them? Well, we can break offense and defense into how they compare when the play type is a run, and when it is a pass. 

### Offense Breakdown: Run vs Pass
Starting with offense and the average EPA gained per play when they throw versus run the ball. 

```{r offense-efficiency, echo=FALSE, warning=FALSE, message=FALSE}

offense <- pbp %>% 
  filter(play_type != "no_play") %>%
  group_by(posteam, play_type) %>%
  summarise(
    epa = mean(epa, na.rm = TRUE),
    frequency = n()
    ) %>%
  pivot_wider(names_from = c("play_type"), values_from = c("epa", "frequency")) %>% 
  dplyr::inner_join(logos, by = c("posteam" = "team_abbr"))

offense %>% 
  ggplot(aes(x = epa_run, y = epa_pass)) +
  ggimage::geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
  ggplot2::geom_hline(aes(yintercept = mean(epa_pass)), color = "red", linetype = "dashed") +
  ggplot2::geom_vline(aes(xintercept = mean(epa_run)), color = "red", linetype = "dashed") + 
  ggplot2::labs(
    x = "Offense Run EPA/play",
    y = "Offense Pass EPA/play",
    caption = "Data: @nflfastR",
    title = "NFL Offensive EPA per Play (Total Season)"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    plot.title = ggplot2::element_text(size = 12, hjust = 0.5, face = "bold")
  ) 
```

Again, this graph has the averages looking at the whole season, what happens when we look at the last three games? 

```{r offense-efficiency-recently, echo=FALSE, warning=FALSE, message=FALSE}
offense <- pbp_recent %>% 
  filter(play_type != "no_play") %>%
  group_by(posteam, play_type) %>%
  summarise(
    epa = mean(epa, na.rm = TRUE),
    frequency = n()
    ) %>%
  pivot_wider(names_from = c("play_type"), values_from = c("epa", "frequency")) %>% 
  dplyr::inner_join(logos, by = c("posteam" = "team_abbr"))

offense %>% 
  ggplot(aes(x = epa_run, y = epa_pass)) +
  ggimage::geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
  ggplot2::geom_hline(aes(yintercept = mean(epa_pass)), color = "red", linetype = "dashed") +
  ggplot2::geom_vline(aes(xintercept = mean(epa_run)), color = "red", linetype = "dashed") + 
  ggplot2::labs(
    x = "Offense Run EPA/play",
    y = "Offense Pass EPA/play",
    caption = "Data: @nflfastR",
    title = "NFL Offensive EPA per Play (Last 2 games)"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    plot.title = ggplot2::element_text(size = 12, hjust = 0.5, face = "bold")
  ) 
```

```{r offense-frequency, echo=FALSE, warning=FALSE, message=FALSE}
offense <- pbp %>% 
  filter( play_type != "no_play") %>%
  group_by(posteam, play_type) %>%
  summarise(
    freq = n()
  ) %>% 
  transmute(
    play_type, percent = freq / sum(freq)
  ) 

offense %>% ggplot(aes(x = reorder(posteam, posteam), y = percent, fill=play_type)) + 
  geom_bar(position=position_fill(reverse = TRUE), stat = "identity") + 
  coord_flip() +
  theme_bw()
  
```


### Defensive Breakdowns: Run vs Pass
Let's do the same thing but looking at how a team's defense fairs against an opposing offense running pass or rush plays. 

```{r defensive-efficiency, echo=FALSE, message=FALSE, warning=FALSE}
defense <- pbp %>% 
  filter(play_type != "no_play") %>%
  group_by(defteam, play_type) %>%
  summarise(
    epa = mean(epa, na.rm = TRUE),
    frequency = n()
    ) %>%
  pivot_wider(names_from = c("play_type"), values_from = c("epa", "frequency")) %>% 
  dplyr::inner_join(logos, by = c("defteam" = "team_abbr"))

defense %>% 
  ggplot(aes(x = epa_run, y = epa_pass)) +
  ggimage::geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
  ggplot2::geom_hline(aes(yintercept = mean(epa_pass)), color = "red", linetype = "dashed") +
  ggplot2::geom_vline(aes(xintercept = mean(epa_run)), color = "red", linetype = "dashed") + 
  ggplot2::labs(
    x = "Defense EPA/play against the Run",
    y = "Defense EPA/play against the Pass",
    caption = "Watch the axes! | Data: @nflfastR",
    title = "NFL Defense EPA per Play (Total Season)"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    plot.title = ggplot2::element_text(size = 12, hjust = 0.5, face = "bold")
  ) + 
  ggplot2::scale_y_reverse() +
  ggplot2::scale_x_reverse()
```

```{r defensive-efficiency-recent, echo=FALSE, message=FALSE, warning=FALSE}
defense <- pbp_recent %>% 
  filter(play_type != "no_play") %>%
  group_by(defteam, play_type) %>%
  summarise(
    epa = mean(epa, na.rm = TRUE),
    frequency = n()
    ) %>%
  pivot_wider(names_from = c("play_type"), values_from = c("epa", "frequency")) %>% 
  dplyr::inner_join(logos, by = c("defteam" = "team_abbr"))

defense %>% 
  ggplot(aes(x = epa_run, y = epa_pass)) +
  ggimage::geom_image(aes(image = team_logo_espn), size = 0.05, asp = 16 / 9) +
  ggplot2::geom_hline(aes(yintercept = mean(epa_pass)), color = "red", linetype = "dashed") +
  ggplot2::geom_vline(aes(xintercept = mean(epa_run)), color = "red", linetype = "dashed") + 
  ggplot2::labs(
    x = "Defense EPA/play against the Run",
    y = "Defense EPA/play against the Pass",
    caption = "Watch the axes! | Data: @nflfastR",
    title = "NFL Defense EPA per Play (Total Season)"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(
    aspect.ratio = 9 / 16,
    plot.title = ggplot2::element_text(size = 12, hjust = 0.5, face = "bold")
  ) + 
  ggplot2::scale_y_reverse() +
  ggplot2::scale_x_reverse()
```

